Stable individual differences in strategies within, but not between, visual search tasks

### 2019-05-28 Manuscript/manuscript.tex
minor typo

\documentclass[Afour, sageh, times]{sagej}
% \RequirePackage{fix-cm}
% \documentclass[a4paper, man, natbib, floatsintext]{apa6}

%% *** Do not adjust lengths that control margins, column widths, etc. ***
\usepackage{subfigure}
% \usepackage{lineno}
% \usepackage{hyperref}
\usepackage{graphicx}
\usepackage{natbib}

@@

As is common in cognitive psychology, most visual search literature has focused on how the average participant performs in the task, despite it being well known that there is a great deal of variability between one subject and the next. From Treisman's work on Feature Integration Theory \citep{treisman1980} to the latest incarnation of the Guided Search Model \citep{wolfe2015}, we have a good understanding of what makes particular objects easier or harder to find. However, these theories and models have neglected the question of why some observers find visual search so much harder than others. These differences can emerge from several different sources of variation: tiredness \citep{mackworth1948}, information-processing ability, speed-accuracy trade-off, motivation, visual impairments \citep{nowakowska2016}, and search strategies \citep{boot2006}. Although their existence has previously been noted \citep{mackworth1948}, a rigorous examination of individual differences in visual search is a challenge that has not been taken up by many researchers, and questions about their impact and stability remain relatively under-explored.

Here we focus on one source of individual differences in visual search: strategy. By strategy we mean a collection of search behaviours from which all observers can freely choose. Examples include adopting a systematic left-to-right and top-to-bottom strategy \citep{gilchrist2006}, or prioritizing locations that, based on knowledge or context, are more likely to contain the target \citep{wolfe2015}. A striking example of the effect of strategy is given by \cite{boot2006}. They asked participants to monitor a cluttered display for an object changing colour or suddenly appearing. Large individual differences were found with respect to the number of saccades participants made while monitoring the stimulus, which was negatively correlated with detection performance.

Eye movement strategies have also been shown to be an important source of individual differences in visual search efficiency. \cite{nowakowska2017} designed a simple search paradigm to discriminate between optimal \citep{najemnik-geisler2008} and stochastic  \citep{clarke2016} search strategies. Participants searched through arrays of line segments (Figure \ref{fig:exampleStimuli}) arranged such that those on one side of the display all had a very similar orientation (homogeneous), while those on the other side had higher variance (heterogeneous). This meant that targets appearing on the homogeneous side were highly salient, while targets on the heterogeneous side were harder to find. The optimal strategy here is to search the heterogeneous half, as targets on the homogeneous side can be detected with peripheral vision. We will refer to this paradigm as the Split-Half Line Segment task (SHLS). Some participants searched the displays near optimally, but others carried out strategies counter to this, failing to even match the performance of the stochastic searcher. The degree to which participants made saccades in line with the optimal search strategy was strongly correlated with the speed of their search. A related version of this paradigm has been used in research investigating eye movement strategies in response to (simulated) hemianopia \citep{nowakowska2016,nowakowska2018}, with similar conclusions: the full spectrum of individual differences in strategy was observed. It is therefore not possible to conclude whether optimal or stochastic models better describe search without first explaining individual variability.

A similar range of strategies, from random to near-optimal, has been found by \cite{irons-leber2016} with the Adaptive Choice Visual Search (ACVS) paradigm. This paradigm involves stimuli made up of small coloured boxes (red, blue, green and a fourth colour that varies from red, through purple, to blue and back again) with numerals written inside them (Figure \ref{fig:exampleStimuli}). The target is a defined as a red or blue box containing one of four numerals (e.g., 2-5), and on each trial one target of each colour is present. The participant's task is to find one of either target as quickly as possible and report the numeral. On trials in which the fourth colour is red (or close to red), participants should search through the blue boxes and report the blue target, as there will be fewer distracters. As the fourth colour changes to through purple to blue, participants should update their strategy and search for the red target. The results showed that participants varied substantially along two key dimensions: how frequently they used the more effective target colour to search (varying from chance performance to near optimal), and how often they changed between colours. Further work \citep{irons-leber2018} has shown that these differences are stable over time (between one and ten days) with test-retest correlations of around $r = 0.83$ ($95\%$ CI $= [0.72, 0.90]$) for optimal choices.

@@

Another example of differences in search strategy comes from the foraging literature  \citep{kristjansson2014,johannesson2016}. In this context, foraging tasks involve searching for multiple targets on each trial. Participants were asked to search through a set of items from four categories, with two categories classed as targets. In the conjunction condition (searching for red-horizontal and green-vertical line segments among red-vertical and green-horizontal distracters), most observers searched in runs, finding all the targets of one target category, and then switching and finding the targets in the other category. This strategy has previously been observed in animal foraging  \citep{dawkins1971}, and suggests holding one complex target template in mind at a time is a better strategy than switching templates. However, a sub-set of participants, termed `super-foragers', were able to change between search target categories with no cost to performance. While test-retest reliability has not been measured explicitly for the foraging paradigm, the task was used as a measure to assess the effect of a six day mindfulness retreat on cognitive performance \citep{hartkamp2017}. From a re-analysis of these data, we can estimate that the test-retest reliability for the mean run length is $r \approx 0.7$ for the feature condition and $r \approx 0.88$ for the conjunction search.

Previous research has investigated the relationship between these behaviours to psychometric measures, but to date, these differences have not shown strong correlations with other attributes. \cite{irons-leber2016,irons-leber2018} found no evidence of a correlation between the proportion of optimal choices made by observers in the ACVS paradigm and measures of visual working memory; trait impulsivity; novelty seeking; need for cognition; and intolerance of uncertainty. Similarly, the differences in foraging behaviour are not accounted for by working memory or inhibitory control \citep{johannesson2017}.

A common theme emerging from these studies is the observation that individual strategies vary in their degree of effectiveness or optimality. However, ``visual search'' encompasses a wide range of tasks, each tapping into a different aspect of behaviour (e.g. feature-based attention, information sampling). The aim of the present study is to investigate the extent to which individual differences are stable across different visual search paradigms. Does it make sense to talk about `super-searchers' who show above-average performance in a range of search tasks (analogous to the 'super-recognizers' of the face-recognition literature  \citep{russell2009})? As a secondary question, we will measure the test-retest reliability of the differences found in the SHLS paradigm, and compare it with existing estimates of reliability for ACVS and MCFT.

@@

The study consists of three paradigms from the visual search literature in which large individual differences have been  found \citep{nowakowska2017, irons-leber2016, kristjansson2014}. Example stimuli can be seen in Figure \ref{fig:exampleStimuli}. A brief overview of each paradigm is given below, with full details in \textit{supplementary materials}. The three tasks were completed over two sessions, approximately one week apart. The SHLS was run in both sessions. The order in which participants completed the tasks was counter-balanced. There are 16 different possible orders of tasks/conditions; four participants completed each order for a total of 64.

The display was presented on a 17-inch CRT monitor with a resolution of $1400 \times 1050$ ($n = 40$) or $1600 \times 1200$ ($n = 24$). Stimulus generation, presentation and data collection were controlled by MATLAB and the psychophysics and eyelink toolboxes  \citep{brainard1997,cornelissen2002} run on a Powermac. Participants sat $\approx 47$cm from the screen.

@@

Our results are consistent with the original SHLS study \citep{nowakowska2017}: we find a large range of individual differences in search reaction time and accuracy (see Figure \ref{fig:splithalf_summary}). These differences are stable across the two sessions, with Pearson's $r \in [0.71, 0.89]$ ($95\%$ CI.) for accuracy in finding hard targets. We get similar scores for the correlation in reaction times between sessions \textit{a} and \textit{b} for hard targets, ($r \in [0.54-0.81]$), easy targets ($r \in [0.52-0.80]$) and target absent trials ($r \in [0.66-0.86]$).

@@

We can also look at the initial search strategies adopted by our participants \ref{fig:splithalf_summary}(c, d). Again, we see large and stable individual differences across the two sessions (test-retest $r \in [0.63, 0.86]$ for the proportion of the first five saccades to the heterogeneous half of the display for target absent trials). More importantly, as with  \citep{nowakowska2017}, we see that the search strategies give a good correlation with reaction times in both session a, $r \in [0.52, 0.82]$ and session b, $r \in [0.50, 0.80]$.

@@

We measured an individual's strategy as the percent of plateau trials in which the individual chose the optimal target (i.e., the target with the fewest distractors: When the variable distractor was red, the optimal choice was blue, and vice versa). The results for the ACVS were consistent with previous findings  \citep{irons-leber2016,irons-leber2018}. We can clearly see from figure \ref{fig:acvs_mcft_summary}(a) that there are individual differences in the proportion of optimal targets reported (range $33.62\%$ - $100.00\%$, $\bar{x} = 59.15$, $s = 16.54$) and the mean ($\log_2$) reaction times (range $1.90$ - $4.80$ seconds). As with the SHLS task, the degree to which participants follow the optimal strategy is correlated with reaction times ($r \in [-0.65, -0.25]$).

@@

The main measure of interest was average run length per trial in the conjunction condition, with a run defined as a succession of one or more of the same target type, which was followed and preceded by the other target or no target. The average run length was the mean number of target selections in a run. The multiple-target foraging results were in line with previous findings  \citep{kristjansson2014,johannesson2016}, with shorter run lengths for feature foraging ($\bar{x} = 3.16$, $s = 3.14$) than conjunction foraging ($\bar{x} = 11.73$, $s = 7.09$). This suggests more frequent foraging for multiple targets concurrently when those targets were defined by features than by conjunctions. Figure \ref{fig:acvs_mcft_summary}(b) depicts the individual differences in terms of run length and the correlation with reaction time ($r \in [-0.55, -0.10]$).

@@

We successfully replicated the wide range of individual differences in strategy and performance that had previously been observed in each of these three visual search paradigms, with a larger sample size than the original experiments. Surprisingly, however, the between-paradigm correlations give $R^2 \approx 0.04$, and this is a generous interpretation that would fail to pass the usual criteria for null hypothesis significance testing. Knowing how one person will behave in one of these paradigms tells us very little about how they will perform in the others. This can be contrasted with the relatively high test-retest correlations of all three of the tasks individually. Indeed, we find that the visual search tasks under consideration here compare favourably to other cognitive psychology paradigms, such as Erkison Flanker and Posner Cueing tasks in terms of test-retest reliability \citep{hedge2018}.

@@

Individual differences pose a challenge for efforts to devise a comprehensive model of visual search. Our understanding of the mechanisms of visual search is based predominantly on experiments that systematically vary details of the search task and measure effects on averaged performance. This approach has led to important insights, for example, about the kinds of visual features that can guide attention (e.g.  \citep{treisman1980, wolfe2007}; how attentional control settings filter  distractors (e.g.  \citep{folk1992,yantis1999}); and biases in attention, such as a bias towards unexplored locations (e.g.  \citep{klein2000}). For all three of the experiments included in the current study, however, the average performance would be highly misleading, as it would describe very few of the individuals' performance. In the original SHLS study, for example, the original aim of the experiment was to assess whether search behaviour could be better described by an optimal  \citep{najemnik-geisler2008}, versus a stochastic  \citep{clarke2016}, model. Considering only the average performance, the stochastic model was a good explanation. Underlying that average performance, however, was a spectrum of search behaviour, replicated here, some of which would be clearly categorized as optimal, and some as stochastic, and some as neither. The original question needed to be refined: for whom is search optimal and for whom is it stochastic? The current findings add even further challenges for researchers, by suggesting we need to account not only for individual differences, but also for the interaction of a given individual with a particular search context.

We view these findings not as a discouraging result, but as thought-provoking and exciting. Vogel and Awh \citep{vogel2008} argued that studying individual differences in cognitive psychology (in their case, working memory) provides valuable insight to constraining potential theories of the underlying cognitive mechanisms. Our results suggest that context and structure of the task also needs to be taken into account. Understanding how an individual's behaviour varies across different search tasks can lead to the development of a comprehensive theory of search.

@@

\bibliographystyle{sageh}
\bibliography{literature}

>>>>

\documentclass[Afour, sageapa, times]{sagej}
% \RequirePackage{fix-cm}
% \documentclass[a4paper, man, natbib, floatsintext]{apa6}

%% *** Do not adjust lengths that control margins, column widths, etc. ***
\usepackage{subfigure}
% \usepackage{lineno}
% \usepackage{hyperref}
\usepackage{graphicx}
% \usepackage{natbib}

@@

As is common in cognitive psychology, most visual search literature has focused on how the average participant performs in the task, despite it being well known that there is a great deal of variability between one subject and the next. From Treisman's work on Feature Integration Theory \cite{treisman1980} to the latest incarnation of the Guided Search Model \cite{wolfe2015}, we have a good understanding of what makes particular objects easier or harder to find. However, these theories and models have neglected the question of why some observers find visual search so much harder than others. These differences can emerge from several different sources of variation: tiredness \cite{mackworth1948}, information-processing ability, speed-accuracy trade-off, motivation, visual impairments \cite{nowakowska2016}, and search strategies \cite{boot2006}. Although their existence has previously been noted \cite{mackworth1948}, a rigorous examination of individual differences in visual search is a challenge that has not been taken up by many researchers, and questions about their impact and stability remain relatively under-explored.

Here we focus on one source of individual differences in visual search: strategy. By strategy we mean a collection of search behaviours from which all observers can freely choose. Examples include adopting a systematic left-to-right and top-to-bottom strategy \cite{gilchrist2006}, or prioritizing locations that, based on knowledge or context, are more likely to contain the target \cite{wolfe2015}. A striking example of the effect of strategy is given by \cite{boot2006}. They asked participants to monitor a cluttered display for an object changing colour or suddenly appearing. Large individual differences were found with respect to the number of saccades participants made while monitoring the stimulus, which was negatively correlated with detection performance.

Eye movement strategies have also been shown to be an important source of individual differences in visual search efficiency. \cite{nowakowska2017} designed a simple search paradigm to discriminate between optimal \cite{najemnik-geisler2008} and stochastic  \cite{clarke2016} search strategies. Participants searched through arrays of line segments (Figure \ref{fig:exampleStimuli}) arranged such that those on one side of the display all had a very similar orientation (homogeneous), while those on the other side had higher variance (heterogeneous). This meant that targets appearing on the homogeneous side were highly salient, while targets on the heterogeneous side were harder to find. The optimal strategy here is to search the heterogeneous half, as targets on the homogeneous side can be detected with peripheral vision. We will refer to this paradigm as the Split-Half Line Segment task (SHLS). Some participants searched the displays near optimally, but others carried out strategies counter to this, failing to even match the performance of the stochastic searcher. The degree to which participants made saccades in line with the optimal search strategy was strongly correlated with the speed of their search. A related version of this paradigm has been used in research investigating eye movement strategies in response to (simulated) hemianopia \cite{nowakowska2016,nowakowska2018}, with similar conclusions: the full spectrum of individual differences in strategy was observed. It is therefore not possible to conclude whether optimal or stochastic models better describe search without first explaining individual variability.

A similar range of strategies, from random to near-optimal, has been found by \cite{irons-leber2016} with the Adaptive Choice Visual Search (ACVS) paradigm. This paradigm involves stimuli made up of small coloured boxes (red, blue, green and a fourth colour that varies from red, through purple, to blue and back again) with numerals written inside them (Figure \ref{fig:exampleStimuli}). The target is a defined as a red or blue box containing one of four numerals (e.g., 2-5), and on each trial one target of each colour is present. The participant's task is to find one of either target as quickly as possible and report the numeral. On trials in which the fourth colour is red (or close to red), participants should search through the blue boxes and report the blue target, as there will be fewer distracters. As the fourth colour changes to through purple to blue, participants should update their strategy and search for the red target. The results showed that participants varied substantially along two key dimensions: how frequently they used the more effective target colour to search (varying from chance performance to near optimal), and how often they changed between colours. Further work \cite{irons-leber2018} has shown that these differences are stable over time (between one and ten days) with test-retest correlations of around $r = 0.83$ ($95\%$ CI $= [0.72, 0.90]$) for optimal choices.

@@

Another example of differences in search strategy comes from the foraging literature  \cite{kristjansson2014,johannesson2016}. In this context, foraging tasks involve searching for multiple targets on each trial. Participants were asked to search through a set of items from four categories, with two categories classed as targets. In the conjunction condition (searching for red-horizontal and green-vertical line segments among red-vertical and green-horizontal distracters), most observers searched in runs, finding all the targets of one target category, and then switching and finding the targets in the other category. This strategy has previously been observed in animal foraging  \cite{dawkins1971}, and suggests holding one complex target template in mind at a time is a better strategy than switching templates. However, a sub-set of participants, termed `super-foragers', were able to change between search target categories with no cost to performance. While test-retest reliability has not been measured explicitly for the foraging paradigm, the task was used as a measure to assess the effect of a six day mindfulness retreat on cognitive performance \cite{hartkamp2017}. From a re-analysis of these data, we can estimate that the test-retest reliability for the mean run length is $r \approx 0.7$ for the feature condition and $r \approx 0.88$ for the conjunction search.

Previous research has investigated the relationship between these behaviours to psychometric measures, but to date, these differences have not shown strong correlations with other attributes. \cite{irons-leber2016,irons-leber2018} found no evidence of a correlation between the proportion of optimal choices made by observers in the ACVS paradigm and measures of visual working memory; trait impulsivity; novelty seeking; need for cognition; and intolerance of uncertainty. Similarly, the differences in foraging behaviour are not accounted for by working memory or inhibitory control \cite{johannesson2017}.

A common theme emerging from these studies is the observation that individual strategies vary in their degree of effectiveness or optimality. However, ``visual search'' encompasses a wide range of tasks, each tapping into a different aspect of behaviour (e.g. feature-based attention, information sampling). The aim of the present study is to investigate the extent to which individual differences are stable across different visual search paradigms. Does it make sense to talk about `super-searchers' who show above-average performance in a range of search tasks (analogous to the 'super-recognizers' of the face-recognition literature  \cite{russell2009})? As a secondary question, we will measure the test-retest reliability of the differences found in the SHLS paradigm, and compare it with existing estimates of reliability for ACVS and MCFT.

@@

The study consists of three paradigms from the visual search literature in which large individual differences have been  found \cite{nowakowska2017, irons-leber2016, kristjansson2014}. Example stimuli can be seen in Figure \ref{fig:exampleStimuli}. A brief overview of each paradigm is given below, with full details in \textit{supplementary materials}. The three tasks were completed over two sessions, approximately one week apart. The SHLS was run in both sessions. The order in which participants completed the tasks was counter-balanced. There are 16 different possible orders of tasks/conditions; four participants completed each order for a total of 64.

The display was presented on a 17-inch CRT monitor with a resolution of $1400 \times 1050$ ($n = 40$) or $1600 \times 1200$ ($n = 24$). Stimulus generation, presentation and data collection were controlled by MATLAB and the psychophysics and eyelink toolboxes  \cite{brainard1997,cornelissen2002} run on a Powermac. Participants sat $\approx 47$cm from the screen.

@@

Our results are consistent with the original SHLS study \cite{nowakowska2017}: we find a large range of individual differences in search reaction time and accuracy (see Figure \ref{fig:splithalf_summary}). These differences are stable across the two sessions, with Pearson's $r \in [0.71, 0.89]$ ($95\%$ CI.) for accuracy in finding hard targets. We get similar scores for the correlation in reaction times between sessions \textit{a} and \textit{b} for hard targets, ($r \in [0.54-0.81]$), easy targets ($r \in [0.52-0.80]$) and target absent trials ($r \in [0.66-0.86]$).

@@

We can also look at the initial search strategies adopted by our participants \ref{fig:splithalf_summary}(c, d). Again, we see large and stable individual differences across the two sessions (test-retest $r \in [0.63, 0.86]$ for the proportion of the first five saccades to the heterogeneous half of the display for target absent trials). More importantly, as with  \cite{nowakowska2017}, we see that the search strategies give a good correlation with reaction times in both session a, $r \in [0.52, 0.82]$ and session b, $r \in [0.50, 0.80]$.

@@

We measured an individual's strategy as the percent of plateau trials in which the individual chose the optimal target (i.e., the target with the fewest distractors: When the variable distractor was red, the optimal choice was blue, and vice versa). The results for the ACVS were consistent with previous findings  \cite{irons-leber2016,irons-leber2018}. We can clearly see from figure \ref{fig:acvs_mcft_summary}(a) that there are individual differences in the proportion of optimal targets reported (range $33.62\%$ - $100.00\%$, $\bar{x} = 59.15$, $s = 16.54$) and the mean ($\log_2$) reaction times (range $1.90$ - $4.80$ seconds). As with the SHLS task, the degree to which participants follow the optimal strategy is correlated with reaction times ($r \in [-0.65, -0.25]$).

@@

The main measure of interest was average run length per trial in the conjunction condition, with a run defined as a succession of one or more of the same target type, which was followed and preceded by the other target or no target. The average run length was the mean number of target selections in a run. The multiple-target foraging results were in line with previous findings  \cite{kristjansson2014,johannesson2016}, with shorter run lengths for feature foraging ($\bar{x} = 3.16$, $s = 3.14$) than conjunction foraging ($\bar{x} = 11.73$, $s = 7.09$). This suggests more frequent foraging for multiple targets concurrently when those targets were defined by features than by conjunctions. Figure \ref{fig:acvs_mcft_summary}(b) depicts the individual differences in terms of run length and the correlation with reaction time ($r \in [-0.55, -0.10]$).

@@

We successfully replicated the wide range of individual differences in strategy and performance that had previously been observed in each of these three visual search paradigms, with a larger sample size than the original experiments. Surprisingly, however, the between-paradigm correlations give $R^2 \approx 0.04$, and this is a generous interpretation that would fail to pass the usual criteria for null hypothesis significance testing. Knowing how one person will behave in one of these paradigms tells us very little about how they will perform in the others. This can be contrasted with the relatively high test-retest correlations of all three of the tasks individually. Indeed, we find that the visual search tasks under consideration here compare favourably to other cognitive psychology paradigms, such as the Erkison Flanker and Posner Cueing tasks in terms of test-retest reliability \cite{hedge2018}.

@@

Individual differences pose a challenge for efforts to devise a comprehensive model of visual search. Our understanding of the mechanisms of visual search is based predominantly on experiments that systematically vary details of the search task and measure effects on averaged performance. This approach has led to important insights, for example, about the kinds of visual features that can guide attention (e.g.  \cite{treisman1980}; how attentional control settings filter  distractors (e.g.  \cite{folk1992,yantis1999}); and biases in attention, such as a bias towards unexplored locations (e.g.  \cite{klein2000}). For all three of the experiments included in the current study, however, the average performance would be highly misleading, as it would describe very few of the individuals' performance. In the original SHLS study, for example, the original aim of the experiment was to assess whether search behaviour could be better described by an optimal  \cite{najemnik-geisler2008}, versus a stochastic  \cite{clarke2016}, model. Considering only the average performance, the stochastic model was a good explanation. Underlying that average performance, however, was a spectrum of search behaviour, replicated here, some of which would be clearly categorized as optimal, and some as stochastic, and some as neither. The original question needed to be refined: for whom is search optimal and for whom is it stochastic? The current findings add even further challenges for researchers, by suggesting we need to account not only for individual differences, but also for the interaction of a given individual with a particular search context.

We view these findings not as a discouraging result, but as thought-provoking and exciting. Vogel and Awh \cite{vogel2008} argued that studying individual differences in cognitive psychology (in their case, working memory) provides valuable insight to constraining potential theories of the underlying cognitive mechanisms. Our results suggest that context and structure of the task also needs to be taken into account. Understanding how an individual's behaviour varies across different search tasks can lead to the development of a comprehensive theory of search.

@@

\bibliographystyle{mslapa}
\bibliography{literature}
